Creating frames of reference for chert exploitation during the Late Pleistocene in Southwesternmost Iberia.

### 2022-12-22 analysis/paper/paper.Rmd
- Pre-knit corrections

Southern and Western Iberia have often been considered key areas to understanding techno-cultural transitions from the Middle Paleolithic to the end of the Upper Paleolithic. As territories located at the western tip of the European continent and with a generally stable climate even during the coldest periods punctuating the Late Pleistocene, these have been regarded as some of the most significant glacial refugia in Europe [@hewitt2000; @gómez2007; @gonzález-sampériz2010; @carvalho2022]. For this reason, Southernwestern Iberia has frequently been at the center of some of the most debated topics regarding Late Pleistocene human adaptations [@finlayson2006; @zilhão2017; @cascalheira2017]. A particularly good example is the region's role in possibly being one of the last territories to be occupied by Neanderthal populations right before their complete disappearance [@finlayson2006; @zilhão2017]. Neanderthals occupied the European continent for more than 300.000 years and are thought to have disappeared while modern humans arrived on the territory [@trinkaus1979; @mellars2004; @finlayson2006; @jennings2011; @rogers2017]. The last territory where Neanderthal populations seemed to exist was Southern Iberia, around c. 37 thousand years ago [@zilhão2017], or perhaps even earlier [@finlayson2006; @tzedakis2007; @carvalho2022]. This region is key to understanding how Neanderthals survived until such a later chronology, the degree and types of interaction those populations may have had with modern humans [@finlayson2006; @zilhão2017], and how and why they eventually went extinct [@dalen2012; @melchionna2018]. Another example is Southwestern Iberia's importance in the discussion of Upper Paleolithic technocomplexes transitions. Previous studies have discussed the territory's potential as a refugium during cold and harsh climatic conditions [@gómez2007; @rodríguez-sánchez2010; @jennings2011]. The Heinrich Event 2 (HE 2) at the onset of the Last Glacial Maximum (LGM), for example, is a period marked by important social and technological transformations. This climatic event was characterized by abrupt and drastic climatic changes that impacted human behavior all across westernmost Europe [@gamble2004; @sanchezgoñi2010]. The identification of a Proto-Solutrean phase in central and southern Portugal with a very distinct index fossil (the Vale Comprido point), and its direct association with the HE 2, put these regions amongst some of the most important case studies of how environmental dynamics have affected human adaptations during the last glacial. Other studies have expanded upon this notion of climatic refugia during harsh climatic events, to understand the Iberian Peninsula as a long-term eco-cultural refugia [@cascalheira2017]. Using this framework, this territory would consist of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources. This continuous use would then create long-term regional adaptive structures, which when correlated with the ecological niches, @cascalheira2017 have referred to as eco-cultural niches. In fact, it seems that a large number of caves and rockshelters in Iberia are multi-layered, giving validity to the aforementioned framework [@schmidt2012]. These eco-cultural niches provide an exceptional opportunity to understand long-term dynamics regarding biotic and abiotic resource exploitation since they can provide details on how human populations maintained or changed their adaptive systems when facing environmental changes, and cultural and social transformations or constraints. One of such possible Late Pleistocene eco-cultural niches to which this theoretical framework has been previously applied is the archaeological site of Vale Boi. This multi-component site is located in westernmost southern Iberia, in a region currently known as the Algarve, and comprises one of the most complete Upper Paleolithic chronocultural sequences of southern Iberia [@bicho2013; @cascalheira2013; @cascalheira2017]. Several cross-scale complex interactions have been identified, displaying resilient behaviors throughout the Upper Paleolithic maintained by their eco-cultural niche, but also adaptation behaviors motivated by niche diversity, social networks, and climatic changes [@cascalheira2017]. Some of these resilient elements are, for example, the continuous use of strategies like grease rendering and selective hunting patterns, site function, certain lithic technology patterns, and the functional specialization of lithic raw materials. The maintenance of social networks through the identification of possible long-distance lithic raw materials has also been suggested [@pereira2016]. The identification of these patterns has been reliant on the great amount of studies that have originated from the archaeological site of Vale Boi [i.e., @bicho2012; @bicho2017; @pereira2016; @manne2012]. A large portion of these studies has focused on lithic technology [@cascalheira2010; @marreiros2015; @belmiro2020], and unlike other regions of Portugal [i.e., @aubry2004; @aubry2009; @aubry2022; @pereira2022; @pereira2021; @costa2022; @jordão2022], archaeological studies focusing on raw materials, and especially chert, have been more scarce, focusing mostly on brief macroscopic results or the differentiation between possibly local and non-local sorts of raw materials. Despite the scarceness of these studies, chert played a significant role in early and later prehistory at Vale Boi [@cascalheira2010; @marreiros2009; @belmiro2020], and across most of Prehistory in Southern Portugal [@zambujo1998; @bicho2003]. As an essential part of late Pleistocene hunter-gatherer adaptations, lithic raw materials have the potential to provide insights into the adaptive strategies of those populations [@féblot-augustins1993; @kuhn1995; @mellars1996], mostly regarding land-use, technological organization, but also cultural and social interactions. In fact, the selection and procurement of raw material has been suggested as a key stage for the technological organization of hunter gatherer groups [@bamforth1997]. Thus, changes in the frequencies of raw materials within the archaeological records may provide evidence for changes in that organization, or even resilience of specific choices, all of which can reflect culturally transmitted preferences within a group of hunter gatherers [@brown1999]. Several models, both formal and informal, have shown that changes of raw material in the archaeological record may reflect different procurement strategies [@binford1985; @binford1979; @gould1985; @gouldsaggers_1985], either related to mobility strategies [@ambrose1990; @kuhn2004; @kuhn1991; @mccall2007] or changes in the availability of raw materials, possibly related to environmental change [@brantingham2003; @brown1999]. Raw materials are also closely linked to the tools produced with them -- the costs of manufacturing technology can be related to the availability of raw materials in the landscape [@torrence1983; @bousman1993; @parry1987; @mackay2011], or the characteristics of the raw materials which might relate to functionality [@mackay2008; @minichillo2006; @stout2002], through the choice of raw materials for specific technologies [@brown1999]. Furthermore, changes in the raw materials can also be related to the establishment of social networks [@whallon2006] and the horizontal transmission of preferences through trade [@brown1999]. In order to explore the aforementioned theories and models to an archaeological site, especially when trying to understand land-use and the adaptive strategies of an eco-cultural niche like what is proposed for southwesternmost Iberia, it is necessary to know the landscape and the raw material sources available in the territory. To distinguish between local and non-local raw materials, first it is necessary to understand the characteristics of the local sources and establish a comparative database which can be used for the analysis of an assemblage. This is true for most of the informal models which try to understand why raw materials change in the archaeological record. Formal models focusing on raw material use posit that it is necessary to have detailed sourcing data in order to adequately apply the models to a region or an archaeological assemblage [@surovell2009; @pop2015]. In fact, the creation of such a database is a starting point for most raw material-focused studies and has been applied all across the world, independent of the studied chronology [i.e., @luedtke1992; @bustillo2009; @driscoll2016; @terradas2017; @ekshtain2022]. In southern Portugal previous efforts have been made to create such a database. The work of @verissimo2004, albeit focusing on the occurrence of chert solely in western Algarve, provided the initial basis for comparative studies with the assemblages from some Late Pleistocene sites based on macroscopic approaches. The creation of LusoLit, a lithotheque currently hosted at the University of Algarve [@pereirab_presenting_2016], and the collection of samples from several outcrops in the region provided a new leap in the study of chert in the region. A few geological studies have also contributed to understand the availability and characteristics of chert in southern Portugal [i.e., @rocha1976; @ribeiro2005]. Nevertheless, these studies are often unpublished or comprehend answers to geological questions, which hamper the comparative use of the data with archaeological assemblages. A large portion of chert-bearing outcrops in southern Portugal remains unstudied, both by archaeologists and geologists. As such, further analysis of the overall variability, location and availability of chert nodules in the Algarve is necessary to start testing behavioural models of land-use and abiotic resource exploitation. Given the potential for a in-depth raw material study at a possible Late Pleistocene eco-cultural niche such as Vale Boi, in this study we explore the location, diversity and availability of chert raw materials in the southernmost region of Portugal -- the Algarve. Our main goal is to establish a reference for chert raw materials in an understudied region in what regards to lithic materials, that can be used for future studies addressing chert exploitation in the Algarve and elsewhere. This includes the development of a methodological approach adapted to the study area, by testing the potential of macroscopic and petrographic approaches for the characterisation of regional cherts. The data presented here is also fully integrated in an online lithotheque (LusoLit) that is now freely available and can be built upon and improved as new data becomes available.

@@

To locate and characterize chert outcrops in southern Portugal and understand chert's characteristics, a macroscopic and petrographic approach was applied to the study of geological samples which were collected through fieldwork. A combination of methods in a multilayered approach is an important tool to reconstruct the geological and geographical origin of raw materials, especially since different methods have their inherent limitations. By combining these two different methodologies the goal was to create several layers of analysis, to which new layers may be added if necessary [@brandl2016]. Several other similar methodologies and approaches have been applied in other regions [i.e., @garcía-simón2016; @terradas2017; @tomasso2019; @desoler2020]. However, the chosen analysis techniques should be adapted to the specific geographic context, the research questions, the problematics, and the characteristics of the types of cherts in question [@luedtke1992]. Since only preliminary studies of raw materials were applied in the western portion of southern Portugal, and petrographic data has been shown to provide good results for the characterization of cherts in this region [@ribeiro2005], the two methodologies were chosen for the study. The geological samples used in this study were obtained during fieldwork, between August 2021 and June 2022. The locations prospected were chosen after reviewing previously known research, which included preliminary raw materials studies in the region [@verissimo2004; @pereira2016], geological scientific papers and theses focusing on the Algarve basin and concerning chert outcrops [@rocha1976; @marques1983; @ramalho1985; @ribeiro2005], and geological maps, which signaled the presence of chert nodules within the outcrops[@rocha1979; @rocha1983; @rocha1989; @oliveira1984; @oliveira1992; @manuppela1987; @manuppella2007]. Unpublished data and coordinates for unsurveyed locations with potential for chert-bearing outcrops gathered during the organization of the LusoLit lithotheque were also checked. Whenever coordinates or specific locations for known outcrops were available, these were directly visited and the surrounding area was surveyed to understand the extension of the outcrops and to locate possible secondary deposition outcrops nearby. When no specific locations were described (for example, in geological maps) several areas with more potential to find chert outcrops within one larger formation were surveyed. Samples were collected whenever possible, focusing on both primary and secondary outcrops. When chert nodules within one single outcrop showed macroscopic differences (such as differences in the color, texture, translucency, or cortex), samples of each different nodule were collected, to cover all chert variability within the outcrop. This variability was also recorded using a database (to distinguish between homogeneous or heterogeneous chert nodules within the outcrop) and through photography (\@ref(fig:fieldphoto)). All samples were registered with resource to a free android app (Archaeosurvey) which was designed for archaeological surveys, and records site location and characteristics [@cascalheira2017a], further adapted for raw material source surveys [@abrunhosa2017]. The version of the software used for fieldwork is an adaption of the latter apps and records data related to outcrop characteristics and conditions (i.e. abundance, visibility, access, geomorphology, chert morphology, and conditions). All data related to the app and fieldwork can be found in the Supplementary Online Materials (SOM). Individual IDs were associated with each sample, which includes sequential numbers (based on recovery order, i.e., SP10) and outcrop code (i.e., PdA).

@@

A two-step approach was applied to characterize the geological samples. First, the samples were analyzed macroscopically following a pre-established dataset. The variables were defined based on specialized literature [@luedtke1992; @crandell2005; @bressy2002], and the dataset with variable description as well as the sample characterizations can be found in the SOM. A small hand lens of 10x magnification was used for this analysis, followed by a higher magnification analysis with resource to a Nikon SMZ25 stereomicroscope, focusing primarily on inclusions and fossil content. Despite several caveats, especially related to the subjectivity and lack of quantitative variables [@bustillo2009], a macroscopic approach is currently still frequently used in chert raw material studies. For a comparative analysis of archaeological artifacts, other methods may be inconvenient or impossible to use, since they may be destructive and often difficult to apply to large assemblages. Macroscopic analyses have the advantage of being less costly and easy to apply. Establishing a reliable macroscopic characterization and understanding the potential of macroscopy to differentiate between cherts, outcrops and formations is essential for comparative studies between geological samples and archaeological assemblages and obtaining preliminary results. By then combining the macroscopic analysis with a petrographic analysis, other studies have shown that the subjective component of this approach can be minimized [@bustillo2009]. The second phase of the study focused on the petrographic analysis of the geological samples. Thin sections were produced from geological samples of all formations, focusing on obtaining petrographic data that reflected the macroscopic variability. In total, 30 thin sections were produced (\@ref(tab:petrolist)), divided into three groups: 1) 20 thin sections of geological samples from different outcrops within the western section of the Algarve; 2) 9 thin sections of geological samples from different outcrops from the eastern section of the Algarve; 3) 1 thin-section of a geological sample recovered from previous works, which was not identified during our survey. Although primary outcrops were prioritized, thin sections of secondary deposition samples were also produced. To compare with the thin sections from this study, other thin sections from previous studies of Jurassic outcrops from western Algarve [@ribeiro2005] were also consulted. All thin sections were analyzed using a Nikon LV100ND or a Leica DM2500 P and following standard petrographic description (full descriptions of the variables considered for the petrographic description can be found in the SOM).

@@

All sample descriptions (macroscopic and petrographic) and accompanying photographs, as well as photographs and data about the outcrops are also available on a database dedicated to the LusoLit lithotheque , which can be accessed online at <https://lusolit.icarehb.com/>. This database will continue to be updated with the other existing samples, non-chert raw materials and future analyses. The complete R code used for all the analysis and visualizations contained in this paper is available at our online research compendium (DOI: 10.17605/OSF.IO/FP7TA). To produce those files, we followed the procedures described by Marwick et al. (2017) for the creation of research compendiums to enhance the reproducibility of research. The files provided contain all the raw data used in our analysis as well as a custom R project [@wickham2015] holding the code to produce all tables and figures. To enable maximum reuse, code is released under the MIT license, data as CC‐0, and figures as CC‐BY (for more information, see [@Marwick2017].

@@

On the eastern part of the Algarve, chert-bearing known formations are from the Middle to Upper Jurassic, known as the Malhão formation and the Jordana formation, respectively. The Malhão formation chert (dating to the Middle Jurassic) was identified in three outcrops. Whenever in a primary outcrop, this chert was homogeneous. The secondary deposits were recent waterlines and slope deposits, and the cherts were often characterized by intense post-depositional alterations. In these cases, it was not possible to confirm the outcrop location. In these outcrops, the nodule frequency varied from common to abundant. The nodules are roundish, ranging between 3 to 5 cm in maximum width. In all cases, access to the outcrops was easy. Although the parent rock was hard, several chert nodules could be collected from the surface, accumulating further down in gentle slope deposits. The Malhão cherts show two differing macroscopic characteristics: pink/reddish cherts and grey cherts. In general, they are both characterized by a dull to medium luster, opaque to sub-translucent translucency, and smooth to semi-smooth feel. They are easily identifiable through the high amounts of macroscopically visible inclusions, which look like white speckling in plain sight. Under the stereomicroscope, several round fossils and long spicule-like shapes can be identified. The petrographic analysis shows for the Malhão cherts from Casal da Colina (primary and secondary outcrops) high amounts of dolomite within the chert. All of these cherts are characterized by a wackestone texture and a high variety of identifiable fossils (although all are poorly preserved and replaced by chalcedony or quartz). These fossils are Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres, and possibly Tentaculites (Fig. \*\*).

@@

The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve (Fig: ). Whenever in a primary outcrop, the chert was homogeneous, although alternated with nodules of other lithologies within the parent rock. No chert was identified in any secondary deposits, which might be related to the anthropic alteration of the landscape. Smaller nodules broken from the parent rock were identified near the primary source in a field. Whenever embedded in the parent rock, the nodules varied in size (\~1-10 cm) and were abundant, with a high level of difficulty in their removal, due to the hardness of the parent rock. The cherts show little macroscopic variability between nodule and outcrop. They are grey/brown (with visible yellow inclusions). Within nodules, however, the cherts are heterogeneous, with dull and shiny or smooth and semi-smooth feel areas. Some of the nodules also show the variability of translucency, with translucent areas, with a very fine grain, and little presence of visible inclusions. The petrographic analysis shows that the cherts range from a wackestone to packstone texture, which was already seen macroscopically. They are composed mostly of microcrystalline quartz, with the presence of fibrous chalcedony replacing the fossils and negligible percentages of other minerals. Fossils are poorly preserved in general, with a few being identifiable: Calcispheres, Bivalve shell, Sponge spicules, Ostracod, Echinoderms, and Gastropod (Fig. \*\*).

@@

It is important to note that the aforementioned chert geological samples represent the chert variability of the identified outcrops. However, a small number of outcrops described in regional geological maps were not identified or were not accessible. This includes four outcrops: 1) a Lower Jurassic outcrop with chert micronodules which was identified through a geological profile (Oliveira, 1992); 2) a Middle Jurassic outcrop located at the easternmost section of the Algarve with partially dolomitized clasts and chert nodules (Oliveira, 1992); 3) several unprospected locations from previous geoarchaeological works, all related to the Peral Anticline and reaching from central to eastern Algarve; 4) a possible chert exploitation archaeological site located on top of a chert source (Monte do Cerro), in eastern Algarve which was not identified and is possibly inaccessible (Bicho et al., 2003). All of these unidentifiable outcrops are located in the central or eastern portion of the Algarve. This may be related to the frequent landscape changes occurring due to agriculture or the population of previously uninhabited areas, which seldom occurs in western Algarve by the cliff areas, where most outcrops are located.

>>>>

Southern and Western Iberia have often been considered key areas to understanding techno-cultural transitions from the Middle Paleolithic to the end of the Upper Paleolithic. As territories located at the western tip of the European continent and with a generally stable climate even during the coldest periods punctuating the Late Pleistocene, these have been regarded as some of the most significant glacial refugia in Europe [@hewitt2000; @gómez2007; @gonzález-sampériz2010; @carvalho2022]. For this reason, Southernwestern Iberia has frequently been at the center of some of the most debated topics regarding Late Pleistocene human adaptations [@finlayson2006; @zilhão2017; @cascalheira2017]. A particularly good example is the region's role in possibly being one of the last territories to be occupied by Neanderthal populations right before their complete disappearance [@finlayson2006; @zilhão2017]. Neanderthals occupied the European continent for more than 300.000 years and are thought to have disappeared while modern humans arrived on the territory [@trinkaus1979; @mellars2004; @finlayson2006; @jennings2011; @rogers2017]. The last territory where Neanderthal populations seemed to exist was Southern Iberia, around c. 37 thousand years ago [@zilhão2017], or perhaps even earlier [@finlayson2006; @tzedakis2007; @carvalho2022]. This region is key to understanding how Neanderthals survived until such a later chronology, the degree and types of interaction those populations may have had with modern humans [@finlayson2006; @zilhão2017], and how and why they eventually went extinct [@dalen2012; @melchionna2018]. Another example is Southwestern Iberia's importance in the discussion of Upper Paleolithic technocomplexes transitions. Previous studies have discussed the territory's potential as a refugium during cold and harsh climatic conditions [@gómez2007; @rodríguez-sánchez2010; @jennings2011]. The Heinrich Event 2 (HE 2) at the onset of the Last Glacial Maximum (LGM), for example, is a period marked by important social and technological transformations. This climatic event was characterized by abrupt and drastic climatic changes that impacted human behavior all across westernmost Europe [@gamble2004; @sanchezgoñi2010]. The identification of a Proto-Solutrean phase in central and southern Portugal with a very distinct index fossil (the Vale Comprido point), and its direct association with the HE 2, put these regions amongst some of the most important case studies of how environmental dynamics have affected human adaptations during the last glacial. Other studies have expanded upon this notion of climatic refugia during harsh climatic events, to understand the Iberian Peninsula as a long-term eco-cultural refugia [@cascalheira2017]. Using this framework, this territory would consist of several ecological niches, consistently used through time, possibly due to the stability in the richness and variety of resources. This continuous use would then create long-term regional adaptive structures, which when correlated with the ecological niches, @cascalheira2017 have referred to as eco-cultural niches. In fact, it seems that a large number of caves and rockshelters in Iberia are multi-layered, giving validity to the aforementioned framework [@schmidt2012]. These eco-cultural niches provide an exceptional opportunity to understand long-term dynamics regarding biotic and abiotic resource exploitation since they can provide details on how human populations maintained or changed their adaptive systems when facing environmental changes, and cultural and social transformations or constraints. One of such possible Late Pleistocene eco-cultural niches to which this theoretical framework has been previously applied is the archaeological site of Vale Boi. This multi-component site is located in westernmost southern Iberia, in a region currently known as the Algarve, and comprises one of the most complete Upper Paleolithic chronocultural sequences of southern Iberia [@bicho2013; @cascalheira2013; @cascalheira2017]. Several cross-scale complex interactions have been identified, displaying resilient behaviors throughout the Upper Paleolithic maintained by their eco-cultural niche, but also adaptation behaviors motivated by niche diversity, social networks, and climatic changes [@cascalheira2017]. Some of these resilient elements are, for example, the continuous use of strategies like grease rendering and selective hunting patterns, site function, certain lithic technology patterns, and the functional specialization of lithic raw materials. The maintenance of social networks through the identification of possible long-distance lithic raw materials has also been suggested [@pereira2016]. The identification of these patterns has been reliant on the great amount of studies that have originated from the archaeological site of Vale Boi [i.e., @bicho2012; @bicho2017; @pereira2016; @manne2012]. A large portion of these studies has focused on lithic technology [@cascalheira2010; @marreiros2015; @belmiro2020], and unlike other regions of Portugal [i.e., @aubry2004; @aubry2009; @aubry2022; @pereira2022; @pereira2021; @costa2022; @jordão2022], archaeological studies focusing on raw materials, and especially chert, have been more scarce, focusing mostly on brief macroscopic results or the differentiation between possibly local and non-local sorts of raw materials. Despite the scarceness of these studies, chert played a significant role in early and later prehistory at Vale Boi [@cascalheira2010; @marreiros2009; @belmiro2020], and across most of Prehistory in Southern Portugal [@zambujo1998; @bicho2003]. As an essential part of late Pleistocene hunter-gatherer adaptations, lithic raw materials have the potential to provide insights into the adaptive strategies of those populations [@féblot-augustins1993; @kuhn1995; @mellars1996], mostly regarding land-use, technological organization, but also cultural and social interactions. In fact, the selection and procurement of raw material has been suggested as a key stage for the technological organization of hunter gatherer groups [@bamforth1997]. Thus, changes in the frequencies of raw materials within the archaeological records may provide evidence for changes in that organization, or even resilience of specific choices, all of which can reflect culturally transmitted preferences within a group of hunter gatherers [@brown1999]. Several models, both formal and informal, have shown that changes of raw material in the archaeological record may reflect different procurement strategies [@binford1985; @binford1979; @gould1985; @gouldsaggers_1985], either related to mobility strategies [@ambrose1990; @kuhn2004; @kuhn1991; @mccall2007] or changes in the availability of raw materials, possibly related to environmental change [@brantingham2003; @brown1999]. Raw materials are also closely linked to the tools produced with them -- the costs of manufacturing technology can be related to the availability of raw materials in the landscape [@torrence1983; @bousman1993; @parry1987; @mackay2011], or the characteristics of the raw materials which might relate to functionality [@mackay2008; @minichillo2006; @stout2002], through the choice of raw materials for specific technologies [@brown1999]. Furthermore, changes in the raw materials can also be related to the establishment of social networks [@whallon2006] and the horizontal transmission of preferences through trade [@brown1999]. In order to explore the aforementioned theories and models in an archaeological site, especially when trying to understand land-use and the adaptive strategies of an eco-cultural niche like what is proposed for southwesternmost Iberia, it is necessary to know the landscape and the raw material sources available in the territory. To distinguish between local and non-local raw materials, first, it is necessary to understand the characteristics of the local sources and establish a comparative database that can be used for the analysis of an assemblage. This is true for most of the informal models which try to understand why raw materials change in the archaeological record. Formal models focusing on raw material use posit that it is necessary to have detailed sourcing data in order to adequately apply the models to a region or an archaeological assemblage [@surovell2009; @pop2015]. In fact, the creation of such a database is a starting point for most raw material-focused studies and has been applied all across the world, independent of the studied chronology [i.e., @luedtke1992; @bustillo2009; @driscoll2016; @terradas2017; @ekshtain2022]. In southern Portugal, previous efforts have been made to create such a database. The work of @verissimo2004, albeit focusing on the occurrence of chert solely in western Algarve, provided the initial basis for comparative studies with the assemblages from some Late Pleistocene sites based on macroscopic approaches. The creation of LusoLit, a lithotheque currently hosted at the University of Algarve [@pereirab_presenting_2016], and the collection of samples from several outcrops in the region provided a new leap in the study of chert in the region. A few geological studies have also contributed to understand the availability and characteristics of chert in southern Portugal [i.e., @rocha1976; @ribeiro2005]. Nevertheless, these studies are often unpublished or comprehend answers to geological questions, which hamper the comparative use of the data with archaeological assemblages. A large portion of chert-bearing outcrops in southern Portugal remains unstudied, both by archaeologists and geologists. As such, further analyses of the overall variability, location, and availability of chert nodules in the Algarve are necessary to start testing behavioral models of land-use and abiotic resource exploitation. Given the potential for a in-depth raw material study at a possible Late Pleistocene eco-cultural niche such as Vale Boi, in this study we explore the location, diversity, and availability of chert raw materials in the southernmost region of Portugal -- the Algarve. Our main goal is to establish a reference for chert raw materials in an understudied region in regards to lithic materials, that can be used for future studies addressing chert exploitation in the Algarve and elsewhere. This includes the development of a methodological approach adapted to the study area, by testing the potential of macroscopic and petrographic approaches for the characterization of regional cherts. The data presented here is also fully integrated into an online lithotheque (LusoLit) that is now freely available and can be built upon and improved as new data becomes available.

@@

To locate and characterize chert outcrops in southern Portugal and understand chert's characteristics, a macroscopic and petrographic approach was applied to the study of geological samples which were collected through fieldwork. A combination of methods in a multilayered approach is an important tool to reconstruct the geological and geographical origin of raw materials, especially since different methods have their inherent limitations. By combining these two different methodologies the goal was to create several layers of analysis, to which new layers may be added if necessary [@brandl2016]. Several other similar methodologies and approaches have been applied in other regions [i.e., @garcía-simón2016; @terradas2017; @tomasso2019; @desoler2020]. However, the chosen analysis techniques should be adapted to the specific geographic context, the research questions, the problematics, and the characteristics of the types of cherts in question [@luedtke1992]. Since only preliminary studies of raw materials were applied in the western portion of southern Portugal, and petrographic data has been shown to provide good results for the characterization of cherts in this region [@ribeiro2005], the two methodologies were chosen for the study. The geological samples used in this study were obtained during fieldwork, between August 2021 and June 2022. The prospected locations were chosen after reviewing previously known research, which included preliminary raw materials studies in the region [@verissimo2004; @pereira2016], geological scientific papers and theses focusing on the Algarve basin and concerning chert outcrops [@rocha1976; @marques1983; @ramalho1985; @ribeiro2005], and geological maps, which signaled the presence of chert nodules within the outcrops[@rocha1979; @rocha1983; @rocha1989; @oliveira1984; @oliveira1992; @manuppela1987; @manuppella2007]. Unpublished data and coordinates for unsurveyed locations with potential for chert-bearing outcrops gathered during the organization of the LusoLit lithotheque were also checked. Whenever coordinates or specific locations for known outcrops were available, these were directly visited and the surrounding area was surveyed to understand the extension of the outcrops and to locate possible secondary deposition outcrops nearby. When no specific locations were described (for example, in geological maps) several areas with more potential to find chert outcrops within one larger formation were surveyed. Samples were collected whenever possible, focusing on both primary and secondary outcrops. When chert nodules within one single outcrop showed macroscopic differences (such as differences in the color, texture, translucency, or cortex), samples of each different nodule were collected, to cover all chert variability within the outcrop. This variability was also recorded using a database (to distinguish between homogeneous or heterogeneous chert nodules within the outcrop) and through photography (\@ref(fig:fieldphoto)). All samples were registered with resource to a free android app (Archaeosurvey) which was designed for archaeological surveys, and records site location and characteristics [@cascalheira2017a], further adapted for raw material source surveys [@abrunhosa2017]. The version of the software used for fieldwork is an adaption of the latter apps and records data related to outcrop characteristics and conditions (i.e. abundance, visibility, access, geomorphology, chert morphology, and conditions). All data related to the app and fieldwork can be found in the Supplementary Online Materials (SOM). Individual IDs were associated with each sample, which includes sequential numbers (based on recovery order, i.e., SP10) and outcrop code (i.e., PdA).

@@

A two-step approach was applied to characterize the geological samples. First, the samples were analyzed macroscopically following a pre-established dataset. The variables were defined based on specialized literature [@luedtke1992; @crandell2005; @bressy2002], and the dataset with the variable descriptions as well as the sample characterizations can be found in the SOM. A small hand lens of 10x magnification was used for this analysis, followed by a higher magnification analysis with resource to a Nikon SMZ25 stereomicroscope, focusing primarily on inclusions and fossil content. Despite several caveats, especially related to the subjectivity and lack of quantitative variables [@bustillo2009], a macroscopic approach is currently still frequently used in chert raw material studies. For a comparative analysis of archaeological artifacts, other methods may be inconvenient or impossible to use, since they may be destructive and often difficult to apply to large assemblages. Macroscopic analyses have the advantage of being less costly and easy to apply. Establishing a reliable macroscopic characterization and understanding the potential of macroscopy to differentiate between cherts, outcrops and formations is essential for comparative studies between geological samples and archaeological assemblages and obtaining preliminary results. By then combining the macroscopic analysis with a petrographic analysis, other studies have shown that the subjective component of this approach can be minimized [@bustillo2009]. The second phase of the study focused on the petrographic analysis of the geological samples. Thin sections were produced from geological samples of all formations, focusing on obtaining petrographic data that reflected the macroscopic variability. In total, 30 thin sections were produced (\@ref(tab:petrolist)), divided into three groups: 1) 20 thin sections of geological samples from different outcrops within the western section of the Algarve; 2) 9 thin sections of geological samples from different outcrops from the eastern section of the Algarve; 3) 1 thin-section of a geological sample recovered from previous works, which was not identified during our survey. Although primary outcrops were prioritized, thin sections of secondary deposition samples were also produced. To compare with the thin sections from this study, other thin sections from previous studies of Jurassic outcrops from western Algarve [@ribeiro2005] were also consulted. All thin sections were analyzed using a Nikon LV100ND or a Leica DM2500 P and following standard petrographic description (full descriptions of the variables considered for the petrographic description can be found in the SOM).

@@

All sample descriptions (macroscopic and petrographic) and accompanying photographs, as well as photographs and data about the outcrops are also available on a database dedicated to the LusoLit lithotheque , which can be accessed online at <https://lusolit.icarehb.com/>. This database will continue to be updated with the other existing samples, non-chert raw materials and future analyses. The complete R code used for all the analysis and visualizations contained in this paper is available at our online research compendium (DOI: 10.17605/OSF.IO/FP7TA). To produce those files, we followed the procedures described by [@Marwick2017] for the creation of research compendiums to enhance the reproducibility of research. The files provided contain all the raw data used in our analysis as well as a custom R project [@wickham2015] holding the code to produce all tables and figures. To enable maximum reuse, code is released under the MIT license, data as CC‐0, and figures as CC‐BY (for more information, see [@Marwick2017].

@@

On the eastern part of the Algarve, chert-bearing known formations are from the Middle to Upper Jurassic, known as the Malhão formation and the Jordana formation, respectively. The Malhão formation chert (dating to the Middle Jurassic) was identified in three outcrops. Whenever in a primary outcrop, this chert was homogeneous. The secondary deposits were recent waterlines and slope deposits, and the cherts were often characterized by intense post-depositional alterations. In these cases, it was not possible to confirm the outcrop location. In these outcrops, the nodule frequency varied from common to abundant. The nodules are roundish, ranging between 3 to 5 cm in maximum width. In all cases, access to the outcrops was easy. Although the parent rock was hard, several chert nodules could be collected from the surface, accumulating further down in gentle slope deposits. The Malhão cherts show two differing macroscopic characteristics: pink/reddish cherts and grey cherts (\@ref(fig:mjplate)). In general, they are both characterized by a dull to medium luster, opaque to sub-translucent translucency, and smooth to semi-smooth feel. They are easily identifiable through the high amounts of macroscopically visible inclusions, which look like white speckling in plain sight. Under the stereomicroscope, several round fossils and long spicule-like shapes can be identified. The petrographic analysis shows for the Malhão cherts from Casal da Colina (primary and secondary outcrops) high amounts of dolomite within the chert. All of these cherts are characterized by a wackestone texture and a high variety of identifiable fossils (although all are poorly preserved and replaced by chalcedony or quartz). These fossils are Sponge spicules, Radiolarians, Ostracods, Echinoderms, Calcispheres, and possibly Tentaculites (\@ref(fig:mjplate)).

@@

The Jordana formation chert (Upper Jurassic) was identified in one area in the Algarve (\@ref(fig:sourcesmap)). Whenever in a primary outcrop, the chert was homogeneous, although alternated with nodules of other lithologies within the parent rock. No chert was identified in any secondary deposits, which might be related to the anthropic alteration of the landscape. Smaller nodules broken from the parent rock were identified near the primary source in a field. Whenever embedded in the parent rock, the nodules varied in size (\~1-10 cm) and were abundant, with a high level of difficulty in their removal, due to the hardness of the parent rock. The cherts show little macroscopic variability between nodule and outcrop. They are grey/brown (with visible yellow inclusions) (\@ref(fig:ujplate)). Within nodules, however, the cherts are heterogeneous, with dull and shiny or smooth and semi-smooth feel areas. Some of the nodules also show the variability of translucency, with translucent areas, with a very fine grain, and little presence of visible inclusions. The petrographic analysis shows that the cherts range from a wackestone to packstone texture, which was already seen macroscopically. They are composed mostly of microcrystalline quartz, with the presence of fibrous chalcedony replacing the fossils and negligible percentages of other minerals. Fossils are poorly preserved in general, with a few being identifiable: Calcispheres, Bivalve shell, Sponge spicules, Ostracod, Echinoderms, and Gastropod (\@ref(fig:ujplate)).

@@

It is important to note that the aforementioned chert geological samples represent the chert variability of the identified outcrops. However, a small number of outcrops described in regional geological maps were not identified or were not accessible. This includes four outcrops: 1) a Lower Jurassic outcrop with chert micronodules which was identified through a geological profile [@oliveira1992]; 2) a Middle Jurassic outcrop located at the easternmost section of the Algarve with partially dolomitized clasts and chert nodules [@oliveira1992]; 3) several unprospected locations from previous geoarchaeological works, all related to the Peral Anticline and reaching from central to eastern Algarve; 4) a possible chert exploitation archaeological site located on top of a chert source (Monte do Cerro), in eastern Algarve which was not identified and is possibly inaccessible [@bicho2003]. All of these unidentifiable outcrops are located in the central or eastern portion of the Algarve. This may be related to the frequent landscape changes occurring due to agriculture or the population of previously uninhabited areas, which seldom occurs in western Algarve by the cliff areas, where most outcrops are located.
